Symmetric Circuits for Rank Gates

### 2018-02-10 sections/transparency-and-unique-extensions.tex
Feb 10

\documentclass[../paper.tex]{subfiles}
\begin{document}
% In contrast with the work of Anderson and Dawar~\cite{AndersonD17}, the
% circuit families we use to establish the main result of this paper are
% additionally required to be transparent. In this subsection we briefly discuss
% this condition, and present results suggesting its necessity.

In the statement of the main theorem of this paper, rather than working with
general symmetric rank-circuits, we instead restrict ourselves to families of
\emph{transparent} circuits. In this section we justify this restriction,
showing first that the a great number of very useful circuit properties (e.g.
syntactic-equivalence) are polynomial-time decidable for transparent circuits
and, second, that deciding most of these circuit properties for general circuits
(with non-symmetric gates) is at least as hard as the graph-isomorphism
problem. Since the proof of the main theorem makes use of the polynomial-time
decidability of many of these properties, these results together justify the
inclusion of the transparency restriction.
% We note additionally that the circuit properties that we show

% In this section we discuss the transparency condition on circuits. It is worth
% noting that, since the symmetric circuits of Anderson and
% Dawar~\cite{AndersonD17} have no non-symmetric internal gates, these circuits
% are transparent. We have from the first part of this subsection that a great
% number of useful properties (and transformations) are polynomial-time
% computable on transparent circuits. However, it is still reasonable to ask if
% the explicit inclusion of the transparency restriction is strictly necessary
% for proving these results, or indeed proving the main result of this paper. We
% provide evidence for necessity of the transparency condition, presenting
% reductions from the graph-isomorphism problem to the problem of deciding some
% of this important properties (e.g. symmetry, syntactic-equivalence, computing
% the orbit of a gate, etc.) over general, not necessarily transparent,
% circuits. We note that in all of these cases the decision problem restricted
% to transparent circuits is polynomial-time decidable.

% We further argue that, since most natural algorithms for transforming an
% arbitrary symmetric circuit into an equivalent transparent circuit make use of
% the polynomial-time decidability of these properties, we should consider these
% hardness results evidence that computing this transformation may be as hard as
% the graph-isomorphism problem
% We also prove that a number of related decision problems (e.g. synatcatic)
% important decision problems are reducible to the graph-isomorphism
% xthat the problem of deciding the orbit of a gate is reduca
% why (or whether) the transparency condition is strictly required when
% considering circuits that contain non-symmetric gates. In the second
% subsection we address this question, proving

% when considering circuits

% It is worth noting that the analogous results of Anderson and Dawar~\ref{}
% make no explicit restriction to transparent circuits,

% mentio It is worth asking if this additional transparency requirement is
% structing It is worth noting that Anderson and Dawar~\ref{} when developing
% their analogous translation, define a polynomial-time algorithm that
% transforms symmetric circuits into equivalent circuits with unique extensions
% -- making no mention of any transparency condition. Of course, the circuits
% they consider contain only symmetric internal gates, and as such are
% transparent already, but we may still ask if the transparency restriction is
% strictly necessary for proving the useful results mentioned above.

% It is worth noting that the translation defined by Anderson and Dawar~\cite{}
% between $P$-uniform families of symmetric circuits (with symmetric gates) and
% formulas of $\FPC$ makes no

% In the second subsection we discuss the necessity of the transparency
% requirement. Anderson and Dawar~\cite{AndersonD17} importance of this trans

% into In the main result of this paper we restrict ourselves to transparent
% circuits, a restriction not found in the analogous result of Anderson and
% Dawar~\cite{}. In this section we develop a number of technical that are
% important for defining and developing the basic technical tools used in this
% paper.

% Anderson and Dawar~\cite{} develop an correspondence between symmetric
% circuits (with symmetric gates) and formulas of fixed point with counting. In
% contrast, in the main theorem of this paper we explicitly require that the
% circuits under discussion be \emph{transparent}. This section we develop a
% number of results about transparent circuits. In the first subsection we
% discuss

% In this section we present a number of useful results for transparent circuits
% that together enable our analysis

% This condition does not appear explicitly in the analogous result of Anderson
% and Dawar~\cite{}, and in this section we discuss the importance of this
% requirement on circuits with non-symmetric gates.

% In the first subsection we demonstrate the usefulness of transparency. We
% begin by showing that we may transform transparent circuits into circuits with
% unique extensions in polynomial-time, allowing us assume, without a loss of
% generality, that $P$-uniform families of transparent circuits also have unique
% labels, and hence (from Proposition~\ref{}) unique extensions. We are also
% able to show that if a circuit has unique labels, which transparency allows us
% to assume, then many natural circuit properties, including the action of a
% permutation on the circuit, as well as the orbit and coarsest supporting
% partition of a gate, are polynomial-time computable.

% These results suggest the importance fundamental to our analysis of circuits.
% The ability to assume that the circuits under discussion have unique
% extensions (and unique labels) allows us to define many important theoretical
% tools, such as a canonical action of $\sym_n$ on the circuit and hence the
% notions of support and supporting partition. Moreover, the polynomial-time
% algorithms for computing the orbit and support of a gate play an important
% role (through the Immerman-Vardi Theorem~\cite{}) in our translation of
% circuits into formulas (see Section \ref{})

% we also show that if a circuit has unique labels (which transparency allows us
% to assume) then there exists polynomial-time algorithms for computing the
% action of $\sym_n$ on the circuit, and the canonical supporting partition and
% the orbit a gate (or elements of a universe of a gate). These algorithms play
% (along with the

% polynomial-time algorithms are thus crucial for developing our translation
% from circuits to formulas.

% and prove the existence of polynomial-time algorithms computing important
% functions, such as the coarsest supporting partition or the orbit of a gate
% (see Proposition~\ref{prop:c} and Lemma~\ref{}).

% Thus the transparency condition enables our analysis of circuits by allowing
% us to assume the circuits also have unique extensions. However, Anderson and
% Dawar~\cite{AndersonD17} prove that symmetric circuit can be transformed into
% circuits with unique extensions,without making any mention of any condition
% analogous to transparency. Of course, our result generalises theirs as they
% only consider circuits with symmetric gates, which are transparent, but it is
% reasonable to ask if the transparency requirement on circuits is strictly
% necessary.

% We address this question in this subsection, presenting reductions
% graph-isomorphism problem to the many related and important problems. We show,
% for example, that there is a polynomial-time many-one reduction from
% graph-isomorphism problem to the problem of deciding the orbit of a gate in an
% arbitrary symmetric circuit, and contrast this result with the known
% polynomial-time algorithm computing orbits of gates in transparent circuits.
% We also present reductions from the graph-isomorphism problem to many related
% problems (e.g. checking syntactic-equivalence or unique label for gates).
% Since most natural algorithms for transforming any an arbitrary symmetric
% circuit into an equivalent family of transparent circuits entail solving these
% related problems, we consider this evidence that computing this transformation
% may be as hard as the graph isomorphism problem.

% Moreover, we show that non-transparent circuits

% an ay be as hard as the graph isomorphism problem. However, we

% show that that being able to transform a circuit into an equivalent
% transparent circuit in polynomial-time would

% make no mention In contrast, Anderson and Dawar~\ref{} do not include any
% specific condition on the circuit families they consider

% consider only symmetric circuits with symmetric gates, which are transparent,
% and as such include no such condition.

% make use of an explicit condition on circuits we call \emph{transparency}. In
% this section we show that transparency is a sufficient condition in that it
% allows us to assume, without a loss of generality, that $P$-uniform families
% of In this subsection we show that if a circuit is transparent then it may be
% transformed in polynomial time into an equivalent circuit with unique labels.
% It follows from Proposition~\ref{prop:unique-labels-unique-extensions}) that
% we may assume, without a loss of generality, that $P$-uniform families of
% transparent circuits have unique labels, and hence unique extensions.

% We also discuss the importance of the transparency condition on circuits
% introduced in this paper. Indeed, while Anderson and Dawar~\cite{AndersonD17},
% place no similar requirement on the symmetric circuits in their paper, in our
% generalisation of their result, transparency is required.

% We also discuss relevant special cases and motivate the explicit requirement
% that the $P$-uniform families of circuit referenced in the main result of this
% paper consist of transparent circuits.

% Anderson and Dawar~\cite{AndersonD17} introduce the notion of a `rigid'
% circuit, a condition analogous to the rigidity condition defined in this
% paper. The following

% \begin{definition}
%   We say a circuit $C_n$ is que labels it has unique extensions. We finally
%   discuss th\emph{rigid} if it has injective labels and for every pair of
%   gates $(g, g')$ in $C_n$ such that $H_g = H_{g'}$ and $g \equiv g'$.
% \end{definition}
\subsection{The Usefulness of Transparency}
In the this subsection we show that a number of important circuit properties are
polynomial-time decidable for transparent circuits (e.g. symmetry, the
syntactic-equivalence relation, etc.). We also provide polynomial-time
algorithms for computing the action of a permutation on a symmetric circuit, as
well as for computing the orbits and supports of gates, and elements of
universes of gates, in a circuit.

@@

% \begin{definition}
%   We say a class of circuits $\mathcal{C}$ is \emph{syntactically transparent}
%   if there is an algorithm that takes in a circuit $C \in \mathcal{C}$ and
%   outputs the syntactic equivalence relation on the gates of $C$ and runs in
%   time polynomial in the size of the input circuit.
% \end{definition}

\begin{lem}
  There is an algorithm that takes as input a transparent circuit $C$ and
  outputs the syntactic-equivalence relation on the gates of $C$. The algorithm
  runs in time polynomial in the size of $C$.
  \label{lem:unique-labels-syntactic-equiv}
\end{lem}
\begin{proof}
  Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  $(\mathbb{B}, \rho)$-circuit of order $n$. We define a relation $\sim$ by
  induction on height (i.e. distance from an input gate). This relation will be
  the syntactic-equivalence relation on $G$. It is easy to check the conditions
  for syntactic-equivalence for gates of height 0 as all such gates are input
  gates and so are either constant gates or relational gates. Suppose $g$ and
  $h$ are gates of height greater than zero. We only need to check conditions
  (i), (iv) and (v) of syntactic-equivalence. It is easy to check (i) and (v).
  If either condition is false then $g$ and $h$ are not syntacticly equivalent.
  So suppose $g$ and $h$ are two internal gates such that $\Sigma (g) =
  \Sigma(h)$ and suppose that if they are both output gates then $\Omega^{-1}(g)
  = \Omega^{-1}(h)$. It remains to check if $L(g)/{\sim}$ and $L(h)/{\sim}$ are
  isomorphism equivalent. First check if $H_h /{\sim} = H_g /{\sim}$. If not,
  $g$ and $h$ are not syntactically equivalent. So suppose $H_h /{\sim} = H_g
  /{\sim}$. If $g$ and $h$ are symmetric gates then $H_h /{\sim} = H_g/{\sim}$
  if, and only if, $L(g)/{\sim}$ and $L(h)/{\sim}$ are isomorphism equivalent,
  so let $g \sim h$. If $g$ and $h$ are not symmetric gates then, since $C$ is
  transparent, $g$ and $h$ have unique labels. We thus have that $L(g)/{\sim}$
  and $L(h)/{\sim}$ are isomorphism equivalent if, and only if, $L(g)^{-1}L(h)$
  acts like an isomorphism. This can be checked by first checking that the
  function induces a sorted permutation on the universe of $g$ (which can be
  checked efficiently). This is sufficient as $L(g)(L(g)^{-1}L(h)) = L(h)$.

  It is easy to see that, given a reasonable binary encoding of a circuit, the
  above procedure can be implemented so as to run in polynomial time, and that
  the relation $\sim$ is the syntactic-equivalence relation on the gates of $C$.

  % Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  % $(\mathbb{B}, \rho)$-circuit of order $n$. We define a relation $\sim$ by
  % induction on height (i.e. distance from an input gate). This relation will be
  % the syntactic-equivalence relation on $G$. It is easy to check the conditions
  % for syntactic-equivalence for gates of height 0 as all such gates are input
  % gates and so are either constant gates or relational gates. Suppose $g$ and
  % $h$ are gates of height greater than zero. We only need to check conditions
  % (i), (iv) and (v) of syntactic-equivalence. It is easy to check (i) and (v).
  % If either condition is false then $g$ and $h$ are not syntacticly equivalent.
  % So suppose $g$ and $h$ are two internal gates such that $\Sigma (g) =
  % \Sigma(h)$ and suppose that if they are both output gates then $\Omega^{-1}(g)
  % = \Omega^{-1}(h)$. It remains to check if $L(g)/{\sim}$ and $L(h)/{\sim}$ are
  % isomorphism equivalent. First check if $H_h /{\sim} = H_g /{\sim}$. If not,
  % $g$ and $h$ are not syntactically equivalent. So suppose $H_h /{\sim} = H_g
  % /{\sim}$. If $g$ and $h$ are symmetric gates then $H_h /{\sim} = H_g/{\sim}$
  % if, and only if, $L(g)/{\sim}$ and $L(h)/{\sim}$ are isomorphism equivalent,
  % so let $g \sim h$. If $g$ and $h$ are not symmetric gates then, since $C$ is
  % transparent, $g$ and $h$ have unique labels. We thus have that $L(g)/{\sim}$
  % and $L(h)/{\sim}$ are isomorphism equivalent if, and only if, $L(g)^{-1}L(h)$
  % acts like an isomorphism. This can be checked by first checking that the
  % function induces a sorted permutation on the universe of $g$ (which can be
  % checked efficiently). This is sufficient as $L(g)(L(g)^{-1}L(h)) = L(h)$.

  % It is easy to see that, given a reasonable binary encoding of a circuit, the
  % above procedure can be implemented so as to run in polynomial time, and that
  % the relation $\sim$ is the syntactic-equivalence relation on the gates of $C$.

\end{proof}

>>>>

\documentclass[../paper.tex]{subfiles}
\begin{document}

In the statement of the main theorem of this paper, rather than working with
general symmetric rank-circuits, we instead restrict ourselves to families of
\emph{transparent} circuits. In this section we justify this restriction,
showing first that the a great number of very useful circuit properties (e.g.
syntactic-equivalence) are polynomial-time decidable for transparent circuits
and, second, that deciding most of these circuit properties for general circuits
(with non-symmetric gates) is at least as hard as the graph-isomorphism
problem. Since the proof of the main theorem makes use of the polynomial-time
decidability of many of these properties, these results together justify the
inclusion of the transparency restriction.

\subsection{The Usefulness of Transparency}
In the this subsection we show that a number of important circuit properties are
polynomial-time decidable for transparent circuits (e.g. symmetry, the
syntactic-equivalence relation, etc.). We also provide polynomial-time
algorithms for computing the action of a permutation on a symmetric circuit, as
well as for computing the orbits and supports of gates, and elements of
universes of gates, in a circuit.

@@

\begin{lem}
  There is an algorithm that takes as input a transparent circuit $C$ and
  outputs the syntactic-equivalence relation on the gates of $C$. The algorithm
  runs in time polynomial in the size of $C$.
  \label{lem:unique-labels-syntactic-equiv}
\end{lem}
\begin{proof}
  Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a transparent
  $(\mathbb{B}, \rho)$-circuit of order $n$. We define the syntactic-equivalence
  relation on $G$ by induction on the height of a gate We note that this
  definition can be implemented by a dynamic program.

  Suppose $g$ and $h$ are gates in $C$ such that $\Sigma (g) = \Sigma(h)$ and
  either both $g$ and $h$ are output gates and $\Omega^{-1}(g) = \Omega^{-1}(h)$
  or neither are output gates. Let $g$ and $h$ be input gates, then set $g
  \equiv h$ if, and only if, both $g$ and $h$ are constant gates or both are
  relational gates and $\Lambda(g) = \Lambda (h)$. Let $g$ and $h$ be internal
  gates and suppose we have defined the syntactic-equivalence relation for all
  gates of height less than the height of either $g$ or $h$. Then $g \equiv h$
  if, and only if, $L(g)$ and $L(h)$ are isomorphism equivalent up to $\equiv$.

  The above algorithm can be implemented by a dynamic program with the number of
  iterations bounded by the number of pairs of gates in $C$, i.e. $\vert C
  \vert$ choose $2$. It remains to show that each iterative step can be
  implemented so as to run in time polynomial in the size of $C$. The only
  condition that is not obviously polynomial-time decidable is checking if
  $L(g)$ is isomorphism-equivalent to $L(h)$ up to $\equiv$. We sketch an
  algorithm that decides this condition in time polynomial in the size of the
  circuit.

  Let $g$ and $h$ be two internal gates in $C$ and suppose we have defined the
  syntactic-equivalence relation for all gates of height less than the height of
  either $g$ or $h$. We test if for each gate in $g' \in H_g$, there exists
  $h_{g'} \in H_h$ such that $g' \equiv h_{g'}$ and $\vert \{g'' \in H_g : g''
  \equiv g'\} \vert = \vert \{h'' \in H_h : h'' \equiv h'\}$. If this test
  fails, we set $g \not\equiv h$.

  We then check if both $g$ and $h$ are symmetric gates or if both are
  non-symmetric gates. Suppose both $g$ and $h$ are symmetric gates. We have
  that $L(g)$ is isomorphism-equivalent to $L(h)$ up to $\equiv$ if, and only
  if, there exists a bijection $f : H_g \rightarrow H_h$ such that $f(g') \equiv
  g'$, for all $g' \in H_g$ if, and only if, for each gate in $g' \in H_g$,
  there exists $h' \in H_h$ such that $g' \equiv h'$ and $\vert \{g'' \in H_g :
  g'' \equiv g'\} \vert = \vert \{h'' \in H_h : h'' \equiv h'\}$. Since we have
  already checked this condition we set $g \equiv h$.

  Suppose $g$ and $h$ are non-symmetric gates. Then, since $C$ is transparent,
  both $g$ and $h$ have unique labels. We thus have that for each $g' \in H_g$
  there is exactly one $h_{g'} \in H_h$ such that $g' \equiv h_{g'}$, and so we
  can define the bijection $f : H_g \rightarrow H_h$ by $f(g') := h_{g'}$ for
  all $g' \in H_g$. It can be shown that $L(g)$ is isomorphism-equivalent to
  $L(h)$ up to $\equiv$ if, and only if, $f \circ L(g)$ is
  isomorphism-equivalent to $L(h)$ if, and only if, $L(h)^{-1} f L(g)$ acts on
  $\ind (g)$ like an element of $\aut(g)$. We have from
  Lemma~\ref{lem:aut-partial} that $\sigma := L(h)^{-1} f L(g)$ acts like an
  element of $\aut(g)$ if, and only if, for all $\vec{a}_R, \vec{b}_{R'} \in \ind(g)$,
  $\sigma$ acts like a partial automorphism on $(\vec{a}_R, \vec{b}_{R'})$. We
  thus conduct the test by iterating through every pair $(\vec{a}_R,
  \vec{b}_{R'}) \in \ind(g)$, and set $g \equiv h$ if, and only if, $\sigma$
  acts like a partial automorphism on every such pair.
\end{proof}
